Scaling up a GF grammar for free parsing

### 2011-12-14 doc/report/report.tex
newer report

\abstract{
This is the report of the work of expanding and enhancing a GF grammar and preparing
it for parsing of free Swedish. 
Outside of this project, an equivalent project for English is being worked upon,
as are tehcniques for making the parser robust.
With this in mind, we have extended the current GF grammar to adapt it for
Swedish specifically so that it can parse an extended set of basic
constructions. We have completed a method for importing the electronical 
lexicon Saldo to a GF format and established a connection to the treebank Talbanken.
The imported lexicon consists of more than 100 000 words
and the grammar now covers ??\% of the xx part of Talbanken. 
% The mapping between the annotations of Talbanken and GF is automatic. 
% The goal is to build a robust parser dealing with uncontrolled natural
%language, a parser which can parse all of Talbanken.

@@

\subsection{A second future tense: ``Kommer att"}
kommer att - for scand, extra tense, example
example table

@@

\subsubsection{advs}
-- or is this really just AdVs? 
to change position of adv:
normal: after verb 'han äpplet redan'
now want: 'äter redan äpplet'
          'har redan ätit äpplet' 
          'har ätit redan äpplet' :( which doesn't work either :)
\subsubsection{Focusing adverbs}
GF separates between two categories of adverbs, the normal ones \verb-Adv-, 
eg. `nu' (\emph{`now'}) and
special ones (name!) \verb-AdV- like `alltid' (\emph{`always'}).
The \verb-AdV- attaches directly to the verb and have a special
field in the \verb-VP--table.
\enumsentence{\shortex{4}
{Jag & äter & aldrig & fisk}
{I & eat & never & fish}
{`I never eat fish'}}
\enumsentence{\shortex{4}
{Jag & äter & fisk& nu }
{I & eat & fish& now }
{`I eat fish now'}}

@@

The kind of noun phrases that can be used seems to coincide with the ones
that can be used with formal subjects (see section `Impersonal constructions').
>> subject controlling verbs
>> why comlicated with doubled types, show simplicity (elegant) with dep.types, and
>> some problems (many fields).
>> samesame as Obj, but conjunction works too, sin is Quant but what i sig?
>> (show circle (sig : NP -> GenNP sig -> sin??)
>> how we can add dependent type for Num to use 'varandra'
>> CN should have dettyp too, 'avståndet till sin bil'

@@

\section{Testing}
Hard do test, can't verify. Regression testing. Talbanken is still too hard.
While developing: regression, 150 sentences to start with. Made longer, both good and bad,
to see how many ways things can be parsed.
Test how many parse trees that are generated.
Look at tables for a VP tex.
Some results.

@@

\chapter{Mapping}
\label{sec:Mapping}
%>> no subjects or object, analyse is more from the lexicon,
%>> compare to others, ask elisabet.
%>> diffs: valencies, pp (gf ingen skillnad direkt/indirekt i träd, pga multiling)
%>> mamba very used and tested for swedish, standard.
%>> focus on what we DO cover.\\
Talbanken contains much valuable information about phrase structures and word 
usage. It is analysed with the MAMBA annotation scheme (Telemann, 1974) which is 
%>> cite mamba? annotation: well intergrated in sw. tradition, descriptive, used
%>> for large scale projects.
%annotation scheme
commonly used, and hence very well-tested, for describing Swedish.
One part of this project has focused translating trees from Talbanken to GF
by constructing a mapping, which 
automatically transforms trees in Talbanken format to GF abstract syntax trees.
We hence get a comparision between the two annotaitons and at the same time
we get methods to extract and translate information to GF notation. 
Figure \ref{fig:translationtrees} shows an example of a visualized Talbanken05 tree
of the sentence \emph{``The cat on the car gets bigger"} and the result of translating it to GF.
Both the POS tags as well as the syntactic information are needed
during the translation, so that all the information shown in the abstract tree
(\ref{pic:gftree}) can be extracted. We need to know that \emph{``katten"} is a noun
used in definite form (\verb-DetArt-), singular (\verb-NumSg-). Even though the
notation is translated, the original analyse from Talbanken
is preserved.
\emph{``På bilen"} should still form a subtree attatched to \emph{``katten"}.
%produces a \verb-NP- from ``katten" this means that
%DetCN : Quant                  -> CN    -> NP
%DetCN  (DetQuant DefArt NumSg)    cat_N = katten

@@

\chapter{Results and discussion}
\label{sec:results}
\section{Evaluation}
%Results from mapping, can probably be improved, bit by bit, depends much on words.
%Valencys would therefor help.  
%Saldo, made a big difference to renew the lexicon. Tested what was missing now from talbanken,
%results. Mostly compound nouns, which we can't expect in the dictionary.
%Is too slow, can't be used for parsing right away
%
%Grammar - hard to evaluate automatically, but Elibet is an expert who has been involved
%in the process to verify the solutions. Testing trees against talbanken.
%The big lexicon makes it very slow, eats all memory
%
%Discussion of the results and methods, why are the results good, why are they not better.
%What other methods could have been used? What did we/I expect and what happend?

>>>>

\abstract{
This is the report of the work of expanding and enhancing a GF grammar and preparing
it for parsing of free Swedish. 
Outside of this project, an equivalent project for English is being worked upon,
as are tehcniques for making the parser robust.
With this in mind, we have extended the current GF grammar to adapt it for
Swedish specifically so that it can parse an extended set of basic
constructions. We have completed a method for importing the electronical 
lexicon Saldo to a GF format and established a connection to the treebank Talbanken.
The imported lexicon consists of more than 100 000 words
and the grammar now covers ??\% of the xx part of Talbanken. 
66\% av easyeasy.
% The mapping between the annotations of Talbanken and GF is automatic. 
% The goal is to build a robust parser dealing with uncontrolled natural
%language, a parser which can parse all of Talbanken.

@@

\subsection{A second future tense: ``Kommer att"}
Swedish and Norwegian have two ways of forming future tense. The standard
form 'ska' is included in the resources as the translation to
'will'. The second tense 'kommer att' is also very commonly used, and
desereved to be covered by our grammar.
Komma is used when we believe that we are sure that it will happen:
'det kommer bli mörkt snart'
and ska more of planning
'jag ska äta mat nu'.
since we believe it will happend, it can not be as in :
'det ska (kommer*) bli fint väder imorn, men de kanske har fel'.

Since two out of the three Scandinavian languages share this tempus, it was 
added to the Scandiniavin extra module.
"jeg kommer til å savne deg"

A table of result can be overview in figure \ref{fig:kommeratt}
\begin{figure}[h]
\begin{verbatim}
s SFutKommer Simul Pos Main : jag kommer att se henne
s SFutKommer Simul Pos Inv  : kommer jag att se henne
s SFutKommer Simul Pos Sub  : jag kommer att se henne
s SFutKommer Simul Neg Main : jag kommer inte att se henne
s SFutKommer Simul Neg Inv  : kommer jag inte att se henne
s SFutKommer Simul Neg Sub  : jag inte kommer att se henne
s SFutKommer Anter Pos Main : jag kommer att ha sett henne
s SFutKommer Anter Pos Inv  : kommer jag att ha sett henne
s SFutKommer Anter Pos Sub  : jag kommer att ha sett henne
s SFutKommer Anter Neg Main : jag kommer inte att ha sett henne
s SFutKommer Anter Neg Inv  : kommer jag inte att ha sett henne
s SFutKommer Anter Neg Sub  : jag inte kommer att ha sett henne
\end{verbatim}
\caption{Parse tree for ``Har han inte ätit de gula äpplena idag".}
\label{gfSwe:parsetree}
\end{figure}

kommer att - for scand, extra tense, example
example table

@@

\subsubsection{advs}
-- or is this really just AdVs? 
-- redan is Adv in abstract. Wrong!!
-- what about 'i regel'?
to change position of adv:
normal: after verb 'han äpplet redan'
now want: 'äter redan äpplet'
          'har redan ätit äpplet' 
          'har ätit redan äpplet' :( which doesn't work either :)
\subsubsection{Focusing adverbs}
GF separates between two categories of adverbs, the normal ones \verb-Adv-, 
eg. `nu' (\emph{`now'}) and
special ones (name!) \verb-AdV- like `alltid' (\emph{`always'}).
The \verb-AdV- attaches directly to the verb and have a special
field in the \verb-VP--table.
\enumsentence{\shortex{4}
{Jag & äter & aldrig & fisk}
{I & eat & never & fish}
{`I never eat fish'}}
\enumsentence{\shortex{4}
{Jag & äter & fisk& nu }
{I & eat & fish& now }
{`I eat fish now'}}

@@

The kind of noun phrases that can be used seems to coincide with the ones
that can be used with formal subjects (see section `Impersonal constructions').
>> subject controlling verbs
>> why comlicated with doubled types, show simplicity (elegant) with dep.types, and
>> some problems (many fields).
>> samesame as Obj, but conjunction works too, sin is Quant but what i sig?
>> (show circle (sig : NP -> GenNP sig -> sin??)
>> how we can add dependent type for Num to use 'varandra'
>> CN should have dettyp too, 'avståndet till sin bil'
\verb-
PhrUtt NoPConj (UttQS (UseQCl (TTAnt TPres ASimul) PPos (QuestVP whatSg_IP (ComplSlash (NPObj ?10 (DetCN ?10 (DetQuant ?10 (IndefArt ?10) NumSg) (UseN woman_N))) (SlashV2a fight_V2))))) NoVoc-
  ajaj unknown here!!

@@

\section{Testing}
For testing has been carried out by using a treebank, consisting of sentences used in 
sub-testsuites. This way we make sure that we doesn't decrease the coverage and will
notice if the grammar starts to overgenerate. Also have some sentences that should not
be parsed. 
When adding new constructions, the table-function is very helpful, shows how a word will
turn out in differnet cases. all tenses, for komma.

@@

\chapter{Extracting a GF treebank} %bootstraping/extracting a GF treebank
\label{sec:Mapping}
%>> no subjects or object, analyse is more from the lexicon,
%>> compare to others, ask elisabet.
%>> diffs: valencies, pp (gf ingen skillnad direkt/indirekt i träd, pga multiling)
%>> mamba very used and tested for swedish, standard.
%>> focus on what we DO cover.\\
Talbanken contains much valuable information about phrase structures and word 
usage. It is analysed with the MAMBA annotation scheme (Telemann, 1974) which is 
%>> cite mamba? annotation: well intergrated in sw. tradition, descriptive, used
%>> for large scale projects.
%annotation scheme
commonly used, and hence very well-tested, for describing Swedish.
One part of this project has focused translating trees from Talbanken to GF
by constructing a mapping, which 
automatically transforms trees in Talbanken format to GF abstract syntax trees.
We hence get a comparision between the two annotaitons and at the same time
we get methods to extract and translate information to GF notation. 
Figure \ref{fig:translationtrees} shows an example of a visualized Talbanken05 tree
of the sentence \emph{``The cat on the car gets bigger"} and the result of translating it to GF.
Both the POS tags as well as the syntactic information are needed
during the translation, so that all the information shown in the abstract tree
(\ref{pic:gftree}) can be extracted. We need to know that \emph{``katten"} is a noun
used in definite form (\verb-DetArt-), singular (\verb-NumSg-). Even though the
notation is translated, the original analyse from Talbanken
is preserved.
\emph{``På bilen"} should still form a subtree attatched to \emph{``katten"}.
%produces a \verb-NP- from ``katten" this means that
%DetCN : Quant                  -> CN    -> NP
%DetCN  (DetQuant DefArt NumSg)    cat_N = katten

@@

\chapter{Results and discussion}
\label{sec:results}
\section{Evaluation}
The grammar in it self is not at all robust, and one unknown word will cause a whole sentence to 
fail. When parsing text, we get low results, for this reason. Not yet interesting to talk about
coverage, but focus on quality and ability to scale up. Our goal standarn is so far Talbanken,
but considering that the low coverage of the grammar without statisticts, a comparison would not be
describing anything, to little indata.
Instead we have the luck to have an expert and her presence has increased the standard.

%Results from mapping, can probably be improved, bit by bit, depends much on words.
%Valencys would therefor help.  
%Saldo, made a big difference to renew the lexicon. Tested what was missing now from talbanken,
%results. Mostly compound nouns, which we can't expect in the dictionary.
%Is too slow, can't be used for parsing right away
%
%Grammar - hard to evaluate automatically, but Elibet is an expert who has been involved
%in the process to verify the solutions. Testing trees against talbanken.
%The big lexicon makes it very slow, eats all memory
%
%Discussion of the results and methods, why are the results good, why are they not better.
%What other methods could have been used? What did we/I expect and what happend?
